---
title: "Data Analysis"
author: "Pavlo Hrab"
date: "4/8/2021"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Read the data 
Firstly, we load the datasets and required libraries

```{r cars, echo=TRUE, message=FALSE, warning=FALSE, paged.print=FALSE}
# Load libraries
library(tidyverse)
library(ggplot2)
library(ggpubr)
library(car)
library(plotly)

# Read the data
df <- as.tibble(read.csv("gapminder_clean.csv", row.names =1))
```

# Walkthrough analysis
## Filter the data, to include only year 1962

```{r pressure, echo=TRUE}
df_1962 <- df %>%
  dplyr::filter(Year==1962)
```

## Plot the correlation
```{r echo=FALSE}
df_1962 %>%
  transmute(percap = gdpPercap, CO2 = CO2.emissions..metric.tons.per.capita.) %>%
  drop_na(percap, CO2) %>%
  ggplot(aes(x = CO2, y=percap)) +
  geom_point() +
  xlab("CO2 emissions") +
  ylab("gdpPercap")
```

The plot above shows a correlation between selected values. Therefore, we are conducting pearson's r calculation for them
```{r}
cor.test(df_1962$gdpPercap, df_1962$CO2.emissions..metric.tons.per.capita., method = "pearson")
```
The calculation suggests, a strong positive correlation between gdpPercap and CO2 emissions with r=0.9260817 and p-value = 2.2e-16.

## Find the correlation across all years


```{r}
# Get all the years (time point) from a dataframe
years <- unique(df$Year)
# Calculate the correlation and store in a list (with years) for all year
res <- lapply(years, function(x){
  filt_df <- df %>%
    filter(Year==x) %>%
    transmute(percap= gdpPercap, co2=CO2.emissions..metric.tons.per.capita.) %>%
    drop_na()
  corr <- cor.test(filt_df$percap, filt_df$co2, method = "pearson")
  list(x,as.numeric(corr$estimate))
})
# Make a vector from a corrs
corrs <- sapply(res, function(x){
  x[[2]]
})
# Find the year, which holds a max pearson's r. 
max_year <- sapply(res, function(x){
  if (x[[2]] == max(corrs) ){
    x[[1]]
  }
})
# Clean NULL elements from an above list. The max year is just a number now
max_year <- as.numeric(max_year[lengths(max_year) != 0] )
```

Therefore we have the max year stored in a variable `max_year` (=1967). 
Then we are constructing an interactive plot, using this year as a filter.

```{r echo=FALSE}
g <- df %>%
  filter(Year==max_year) %>%
  transmute(percap=gdpPercap, co2=CO2.emissions..metric.tons.per.capita., pop=pop, continent=continent) %>%
  drop_na(percap, co2, pop, continent) %>%
  filter(continent!="") %>%
  ggplot(aes(y=percap, x=co2)) +
  geom_point(aes(size=pop, color=continent)) +
  xlab("CO2 emissions") +
  ylab("gdpPerCap") 
ggplotly(g)
```

# Exploratory data analysis without an instruction
## Relationship between `continent` and `Energy use` variables
The first thing, first - we should plot the data to get the know it

```{r echo=FALSE}
g <- df %>%
  transmute(continent=continent, energy=Energy.use..kg.of.oil.equivalent.per.capita.) %>%
  drop_na() %>%
  filter(continent != "") %>%
  ggplot(aes(x=continent, y=energy, fill=continent)) +
  geom_boxplot()+
  geom_jitter(color="black", size=0.2, alpha=0.7) +
  ggtitle("Energy use across continents from 1962 till 2007") +
  xlab("Continent") +
  ylab("Energy use")
ggplotly(g)
```

### Stats
The plot suggests that we have one categorical variable and one quantitive The best statistical test for this kind of problem is ANOVA test. 
However there are several assumptions, we need to secure before applying test
1. Observations obtained independly and randomly
2. The data of each factor level is normally distributed
3. These populations have common variance

The first assumption is secured, we assume that these variables obtained independently and randomly. For the rest, we should run some tests

Let's filter the data for the tests
```{r}
df_energy <- df %>%
  transmute(continent=continent, energy=Energy.use..kg.of.oil.equivalent.per.capita.) %>%
  drop_na() %>%
  filter(continent != "")
```


#### Check the variance homogenuity
To test this assumption, we would use Levene's test from car package, because it's less sensitive to departures from normal distribution, than the analogous Bartlett’s test
```{r}
leveneTest(energy ~ continent, data=df_energy)
```
The null hypothesis in Levene's test is that groups have common variance. We just broke it with p-value 8e-10. Therefore we cannot assume that the variance is common between population

### Check if population have normal distribution
We should also check if the data is normally distributed between continents. For that we should use Shapiro-Wilk test. 

```{r}
continents <-  unique(df_energy %>% select(continent))$continent

res <- lapply(continents, function(x){
  df_tmp <- df_energy %>%
    filter(continent==x)
  list(shapiro.test(df_tmp$energy), x)
  })

for (i in res){
  print(paste(i[[2]], "have W: ", i[[1]]$statistic, " with p-value ", i[[1]]$p.value))
}
```
The data above suggest, that the normality assumption is violated. However, the Shapiro-Wilk test tends to pinpoint the slightest normality violation, if the data is big (>50 variables). Therefore, it always better to check with QQ-plots
```{r echo=FALSE}
ggqqplot(df_energy, x = "energy", facet.by = "continent")
```

The QQ-plots aligns with the results of the shapiro-wilk test. The only data for the Oceania have normal distribution.
Just to be sure, let's print the stats for each continent
```{r}
group_by(df_energy, continent) %>%
  summarise(
    count = n(),
    mean = mean(energy, na.rm = TRUE),
    median=median(energy, na.rm = TRUE),
    sd = sd(energy, na.rm = TRUE)
  )
```
The small amount of data partly explains the Oceania normality. Given that the Energy use grow with the passage of time, we expect non-normal distribution across 10 years.

So, we should non-parametric tests instead. The analogous to ANOVA non-parametric test is Kruskal-Wallis test
```{r}
kruskal.test(energy ~ continent, data = df_energy)
```
The Kruskal-wallis test us, that there is an statistically significant difference in medians with the populations. However we haven't compared them pairwise. Because the normality assumption for 4 out of 5 groups is violated we could use pairwise comparisons with Wilcoxon rank sum test.

```{r}
pairwise.wilcox.test(df_energy$energy, df_energy$continent,
                 p.adjust.method = "BH")
```
The results above suggest, that there is a significant difference in energy use over the 10 time points for 5 continents. The only not significant difference with p-value of 0.09746 is between Asia and Americas.

Let's plot the final plot with all the stats included
```{r echo=FALSE}
my_comparisons <-  combn( continents , 2 )

df %>%
  transmute(continent=continent, energy=Energy.use..kg.of.oil.equivalent.per.capita.) %>%
  drop_na() %>%
  filter(continent != "") %>%
  ggplot(aes(x=continent, y=energy, fill=continent)) +
  geom_boxplot()+
  geom_jitter(color="black", size=0.2, alpha=0.7) +
  ggtitle("Energy use across continents from 1962 till 2007") +
  xlab("Continent") +
  ylab("Energy use") +
  stat_compare_means(method = "kruskal.test", label.x = 1, label.y = 31000) +
  stat_compare_means(comparisons = as.list(as.data.frame(my_comparisons)))
```


##Imports of goods and services analysis

The point of this analysis is to check whether there is a significant difference in terms of imports between Europe and Asia in years after 1990
Let's filter the data first
```{r}
df_1990 <- df %>%
  filter(Year>1990) %>%
  transmute(continent=continent,import=Imports.of.goods.and.services....of.GDP. ) %>%
  filter(continent=="Asia" | continent=="Europe") %>%
  drop_na()
```

Let's create a simple boxplot to get the sense of a data

```{r echo=FALSE}
g <- df_1990 %>%
  ggplot(aes(x=continent, y=import, fill=continent)) +
  geom_boxplot()+
  geom_jitter(color="black", size=0.2, alpha=0.7) +
  ggtitle("Energy use across continents from 1990 till 2007") +
  xlab("Continent") +
  ylab("Energy use")
ggplotly(g)
```

We have two groups, and we check the difference between them. We cannot use one-sample t-test, because the groups are independent, there are no reference to compare one of them. Therefore, the Unpaired Two-Samples T-test is a better option. But we first need to meet the assumption for this parametric test:
1. Normal distribution within groups
2. Variance in these groups is equal

Let's perform the normality check using shapiro-wilk test and using QQ-plot
```{r}
continents <-  unique(df_1990 %>% select(continent))$continent

res <- lapply(continents, function(x){
  df_tmp <- df_1990 %>%
    filter(continent==x)
  list(shapiro.test(df_tmp$import), x)
  })

for (i in res){
  print(paste(i[[2]], "have W: ", i[[1]]$statistic, " with p-value ", i[[1]]$p.value))
}
```
```{r echo=FALSE}
ggqqplot(df_1990, x = "import", facet.by = "continent")
```
And let's explore the basic stats for each group
```{r}
group_by(df_1990, continent) %>%
  summarise(
    count = n(),
    mean = mean(import, na.rm = TRUE),
    median=median(import, na.rm = TRUE),
    sd = sd(import, na.rm = TRUE)
  )
```

We can see, that the normality assumption is violated. Shapiro-Wilk test gave the p<0.05 for each group. Given the sensitivity of this test, we plotted the QQ-plot, which, unfortunately, confirms the violation of this assumption.

Then let's If the variances are equal between groups 
```{r}
var.test(import ~ continent, data = df_1990)
```
As we see, the homogeneity of the variances rule is also violated (p-value<<0.05)

So, then we should use the non-parametric alternative, which is Unpaired Two-Samples Wilcoxon Test.

```{r}
wilcox.test(import~continent,data=df_1990, alternative = "two.sided")
```
Well, the difference between two groups is not statistically significant (p-value=0.7867).

And the final visualization^

```{r echo=FALSE}
df_1990 %>%
  ggplot(aes(x=continent, y=import, fill=continent)) +
  geom_boxplot()+
  geom_jitter(color="black", size=0.2, alpha=0.7) +
  ggtitle("Energy use across continents from 1990 till 2007") +
  xlab("Continent") +
  ylab("Energy use")+
  stat_compare_means(comparisons = list(c("Asia", "Europe")))
```


## Highest Population density across coutries

Let's transform the initial data, to save only population density, country and year

```{r}
df_pop <- df%>%
  transmute(year=Year, country=Country.Name, pop_den=Population.density..people.per.sq..km.of.land.area.) %>%
  drop_na()
```


Let's take a look at a data with an interactive plot.

```{r echo=FALSE}
g <- df_pop %>%
  ggplot(aes(x=year, y=pop_den, group=country))+
  geom_line() +
  xlab("Years") +
  ylab("Population density")+
  ggtitle("Population density over years for countries")
ggplotly(g)
```

We could tell right away, that the highest population density across all years is `Macao SAR, China` or `Monaco`. However, we cannot tell which a winner across all years.



```{r}
# Get unique counties and years to a vectors
countries <- unique(df_pop$country)
years <- unique(df_pop$year)
# Make an empty dataframe with countries, which we will populate
coutry_rank <- as.data.frame(unique(df_pop$country))
colnames(coutry_rank) <- "country"
# Iterate over years, rank the population density, add ranks to a dataframe
for (i in years){
  df_tmp <- df_pop %>%
    filter(year==i) 
  
  order(df_tmp$pop_den)
  data_ranked <- as.data.frame(cbind(df_tmp$country, as.numeric(rank(df_tmp$pop_den))))
  coutry_rank <- cbind(coutry_rank, as.numeric(data_ranked[match(coutry_rank$country, data_ranked$V1), ]$V2))
}
# Rename columns
colnames(coutry_rank) <- c("country", years)
# Construct a matrix out of dataframe only with numbers (to compute mean)
rank_matrix <- coutry_rank %>%
  select(-country) %>%
  as.matrix()
coutry_rank <- cbind(coutry_rank, rowMeans(rank_matrix))
coutry_rank[is.na(coutry_rank)] = 0
colnames(coutry_rank) <- c("country", years, 'average')

# Print the rows of a dataframe with the highest rank (the bigger, the better)
coutry_rank %>%
  filter(average==max(average))
```

So, then we conclude, that the countries, that have the highest population density over years are `Macao SAR, China` and `Monaco`.

## Life expectancy increase

At first filter the data. Select years, life expectancy, country
```{r}
df_life <- df %>%
  transmute(year=Year, country=Country.Name, life=Life.expectancy.at.birth..total..years.) %>%
  drop_na()

```

Look at the data.
```{r echo=FALSE}
g <- df_life %>%
  ggplot(aes(x=year, y=life, group=country))+
  geom_line() +
  xlab("Years") +
  ylab("Life expectancy")+
  ggtitle("Life expectancy over years for countries")
ggplotly(g)
```

The plot is less than informative. We should pinpoint the greatest increase in life expectancy since 1962. In other words, the greatest difference between 2007 and 1962. Let's compute it
```{r}
df_life <- df_life %>%
  pivot_wider(names_from = year, values_from=life)

matrx <- df_life %>%
  select(-country)

res <- apply(matrx, 1, function(x){
  lst <- x[!is.na(x)]
  diffs <- lst[length(lst)]-lst[1]
  diffs
})

df_life$diff <- res
top_diffs <- filter(df_life, diff %in% sort(df_life$diff, decreasing = T)[1:10])[c(1,2,11,12)]
top_diffs
```

Lets make the plot one more time, but color the top countries this time

```{r echo=FALSE}
df_life <- df %>%
  transmute(year=Year, country=Country.Name, life=Life.expectancy.at.birth..total..years.) %>%
  drop_na()
g <- df_life %>%
  mutate(color=ifelse(country %in% top_diffs$country,1,0)) %>%
  ggplot(aes(x=year, y=life, group=country, color=color))+
  geom_line() +
  xlab("Years") +
  ylab("Life expectancy")+
  ggtitle("Life expectancy over years for top rating countries")
ggplotly(g)
```





